TBD

### 2018-10-19 paper/paper.tex
reworked some stuff

\section{introduction}\label{sec:intro}
	Gaussian processes have been used extensively in astrophysics to model stochastic variability across space and time. 
	They have been used to study variable stars and AGN, to model the extragalactic distribution of dust, and to model correlated noise in 
	photometric and radial velocity observations of exoplanet host stars.	\gordontodo{X-ray binaries?}

	\gordontodo{The following paragraph might make more sense in the exomoons section.}
	Gaussian process noise models are most advantageous when correlated noise dominates over white noise. 
	This is already the case for state-of-the-art radial velocity observations, and transit 
	photometry is not far behind. Future space-based observatories such as James Webb 
	will enable high-precision photometry in which stellar variability is the main barrier to detecting ever more shallow transit signals. 
	Stellar variability which obscures transit signal can be a result of spot and faculae crossings, asteroseismic oscillations, or variable 
	granulation on the star's surface. GPs are able to \gordontodo{something about how a GP model is non-parametric and the advantages thereof? Also agnostic 
	to the source of variability (seems related)}

@@

\section{gaussian processes and the celerite model}
	A Gaussian process is a stochastic noise model which consists of a mean:
	\begin{equation} 
		\bvec{\mu_\theta}(\bvec{x}) = \expandvec{\mu_\bvec{\theta}(x_1)}{\mu_\bvec{\theta}(x_N)}
	\end{equation}
	and covariance defined the kernel function $k_\bvec{\alpha}(x_n, x_m)$. The mean and kernel functions depend on the parameters 
	$\bvec{\theta}$ and $\bvec{\alpha}$ respectively, which are referred to as ``hyperparameters" of the GP. The kernel function defines a covariance matrix 
	$\left[K\right]_{n, m} = k_\alpha(x_n, x_m)$. This model has a likelihood function given by
	\begin{equation}
		\ln\ \mathcal{L(\bvec{\theta}, \bvec{\alpha})} = \ln\ p(\bvec{y}|X, \bvec{\theta}, \bvec{\alpha}) = 
			-\frac{1}{2}\bvec{r_\theta}^\T K_\bvec{\alpha}^{-1}\bvec{r_\theta} 
			-\frac{1}{2}\ln\ \mathrm{det}(K_\bvec{\alpha}) - \frac{N}{2}\ln(2\pi)
	\end{equation}
	\gordontodo{I'm using $X$ here because DFM does, but wouldn't $\bvec{x}$ be more clear?}
	where 
	\begin{equation}
		\bvec{y} = \expandvec{y_1}{y_N}
	\end{equation}
	are the data, $N$ is the number of datapoints, and
	\begin{equation}
		\bvec{r_\theta} = \bvec{y}-\bvec{\mu_\theta}(\bvec{x})
	\end{equation}
	By maximizing the likelihood (or, in practice, minimizing the negative of the log-likelihood), one can obtain an estimate of the hyperparameters $\bvec{\theta}$ and 
	$\bvec{\alpha}$. Furthermore, Markov Chain Monte Carlo (MCMC) methods can be used to sample the likelihood in hyperparameter-space. \gordontodo{expand 
	on this.}
	\subsection{The \celerite model}
		The primary limitation of Gaussian process noise models is that they are computationally expensive, with computation of the likelihood function scaling as 
		$\mathcal{O}(N^3)$ due to the matrix inversion. This is especially problematic for use cases such as MCMC which require calling the likelihood function many 
		times. This has prevented the adoption of Gaussian process noise models for datasets larger than $N \approx 10^3$, except in cases where approximate 
		methods suffice. The \celerite method overcomes this issue by reducing the time to compute the likelihood function to $\mathcal{O}(NJ^2)$ for one 
		class of kernel functions; where J is the number of \celerite terms making up the kernel function of the GP. This class of kernel functions are given by 
		\begin{equation}
			k_\alpha(t_n, t_m) = \sigma_n^2 \delta_{nm} + \sum_{j=1}^J a_j e^{-c_j\tau_{nm}}
		\end{equation}
		where $\bvec{\alpha} = (a_1...a_j, c_1...c_j)$, $\sigma_n^2$ is the variance of the white noise, and $\tau_{nm} = |t_n-t_m|$. 
		The coefficients $a$ and $c$ may be complex.

@@

	\subsection{sampling the distribution}
	A gaussian process can be used to simulate correlated noise. If $L$ is the lower triangular factor from the Cholesky decomposition of the 
	$(N\times N)$ covariance matrix $K$, and $\bvec{z}$ is a vector of length $N$ containing random numbers drawn for a standard normal distribution, i.e.
	\begin{equation}
		\bvec{z} \sim \mathcal{N}(1, 0)
	\end{equation}
	then $\bvec{x} = \bvec{\mu_x} + L\bvec{z}$ is a vector with entries correlated by $K$ and with mean $\bvec{\mu_x}$, i.e.
	\begin{equation}
		\bvec{x} \sim \mathcal{N}(\bvec{\mu_x}, K)
	\end{equation}
	\gordontodo{fix this so that $\mu$ depends on $\theta$ and is a function of $\bvec{t}$ or something like that. Compuational cost should be $\mathcal{O}(NJM^2)$}

@@

\section{example: exomoons}
	\gordontodo{What if I used Kepler-1625 as an example? Could simulate WFC3 GRISM data and show that moon could be validated with Celerite2D?}

>>>>

\section{introduction}\label{sec:intro}
	A Gaussian process is a generalization of the Gaussian distribution to functions \citep{Rasmussen:2006}. They have been used 
	extensively in astrophysics as a model of stochastic variability across both space and time to study variable stars, AGN, the extragalactic 
	dust distribution, and to model correlated noise in photometric and radial velocity measurements of exoplanet hosts. \gordontodo{X-ray binaries?}

	\gordontodo{something about how a GP model is non-parametric and the advantages thereof? Also agnostic 
	to the source of variability (seems related) --- essentially: what makes a GP noise model desirable over alternatives?}

	Unfortunately, most GP operations are computationally expensive. The common tasks of computing a likelihood function, sampling from a GP, and 
	interpolating or smoothing data all run in $\mathcal{O}(N^3)$ time for $N$ data points \gordontodo{make sure this is true for interpolation}. This is 
	especially problematic for use cases that require iterative calls to the likelihood function, as is the case when minimizing the likelihood or performing 
	Markov Chain Monte Carlo simulations. This has prevented the adoption of GP methods for datasets larger than $N \sim 10^3$.

	The issue of computational expense can be partially overcome either by approximating a GP (often by exploiting the properties of sparse 
	matrices), or by restricting the user to a subset of GPs for which the computation can be sped up. One member of the latter class is the 
	\celerite method, which increases the speed of most common GP computations to $\mathcal{O}(NJ^2)$ where $J$ is typically small compared 
	to $N$. \celerite achieves this increased speed by limiting the user to GPs with a specific functional form of the covariance. Additionally, \celerite 
	only allows for one-dimensional GPs.

@@

\section{gaussian processes and the celerite model}
	A Gaussian process is a stochastic noise model which consists of a mean:
	\begin{equation} 
		\bvec{\mu_\theta}(\bvec{x}) = \expandvec{\mu_\bvec{\theta}(x_1)}{\mu_\bvec{\theta}(x_N)}
	\end{equation}
	and covariance defined the kernel function $k_\bvec{\alpha}(x_n, x_m)$. The mean and kernel functions depend on the parameters 
	$\bvec{\theta}$ and $\bvec{\alpha}$ respectively, which are referred to as ``hyperparameters" of the GP. The kernel function defines a covariance matrix 
	$\left[K\right]_{n, m} = k_\alpha(x_n, x_m)$. This model has a likelihood function given by
	\begin{equation}
		\ln\ \mathcal{L(\bvec{\theta}, \bvec{\alpha})} = \ln\ p(\bvec{y}|X, \bvec{\theta}, \bvec{\alpha}) = 
			-\frac{1}{2}\bvec{r_\theta}^\T K_\bvec{\alpha}^{-1}\bvec{r_\theta} 
			-\frac{1}{2}\ln\ \mathrm{det}(K_\bvec{\alpha}) - \frac{N}{2}\ln(2\pi)
	\end{equation}
	\gordontodo{I'm using $X$ here because DFM does, but wouldn't $\bvec{x}$ be more clear?}
	where 
	\begin{equation}
		\bvec{y} = \expandvec{y_1}{y_N}
	\end{equation}
	are the data, $N$ is the number of datapoints, and
	\begin{equation}
		\bvec{r_\theta} = \bvec{y}-\bvec{\mu_\theta}(\bvec{x})
	\end{equation}
	By maximizing the likelihood (or, in practice, minimizing the negative of the log-likelihood), one can obtain an estimate of the hyperparameters $\bvec{\theta}$ and 
	$\bvec{\alpha}$. Furthermore, Markov Chain Monte Carlo (MCMC) methods can be used to sample the likelihood in hyperparameter-space. \gordontodo{expand 
	on this.}
	\subsection{The \celerite model}
		The primary limitation of Gaussian process noise models is that they are computationally expensive, with computation of the likelihood function scaling as 
		$\mathcal{O}(N^3)$ due to the matrix inversion. This is especially problematic for use cases such as MCMC which require calling the likelihood function many 
		times. This has prevented the adoption of Gaussian process noise models for datasets larger than $N \approx 10^3$, except in cases where approximate 
		methods suffice.

		The \celerite method overcomes this issue by reducing the time to compute the likelihood function to $\mathcal{O}(NJ^2)$ for one 
		class of kernel functions; where J is the number of \celerite terms making up the kernel function of the GP. This class of kernel functions are given by 
		\begin{equation}
			k_\alpha(t_n, t_m) = \sigma_n^2 \delta_{nm} + \sum_{j=1}^J a_j e^{-c_j\tau_{nm}}
		\end{equation}
		where $\bvec{\alpha} = (a_1...a_j, c_1...c_j)$, $\sigma_n^2$ is the variance of the white noise, and $\tau_{nm} = |t_n-t_m|$. 
		The coefficients $a$ and $c$ may be complex.

@@

	\subsection{sampling the distribution}
	A gaussian process can be used to simulate correlated noise. If $L$ is the lower triangular factor from the Cholesky decomposition of the 
	$(N\times N)$ covariance matrix $K$, and $\bvec{z}$ is a vector of length $N$ containing random numbers drawn for a standard normal distribution, i.e.
	\begin{equation}
		\bvec{z} \sim \mathcal{N}(1, 0)
	\end{equation}
	then $\bvec{x} = \bvec{\mu_\bvec{\theta}(\bvec{x})} + L\bvec{z}$ is a vector with entries correlated by $K$ and with mean $\bvec{\mu_\theta}$, i.e.
	\begin{equation}
		\bvec{x} \sim \mathcal{N}(\bvec{\mu_\theta}, K)
	\end{equation}
	\gordontodo{fix this so that $\mu$ depends on $\theta$ and is a function of $\bvec{t}$ or something like that. Compuational cost should be $\mathcal{O}(NJM^2)$}

@@

\section{example: exomoons}
	\gordontodo{What if I used Kepler-1625 as an example? Could simulate WFC3 GRISM data and show that moon could be validated with Celerite2D?}
	Gaussian process noise models are most advantageous when correlated noise dominates over white noise. 
	This is already the case for state-of-the-art radial velocity observations, and transit 
	photometry is not far behind. Future space-based observatories such as James Webb 
	will enable high-precision photometry in which stellar variability is the main barrier to detecting ever more shallow transit signals. 
	Stellar variability which obscures transit signal can be a result of spot and faculae crossings, asteroseismic oscillations, or variable 
	granulation on the star's surface.
